---
title: "Tidy Tuesday NHL"
author: "Anatoly Tsyplenkov"
date: '2020-03-03'
categories: [R, tidytuesday]
code-fold: true
---

Okay, here it is. My first contribution to the my website. I'll try to make a very quick and simple data exploration analysis for this weeks' [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-03/readme.md). There are three datasets available. I'm not sure that I'll be able to analyze all of them… Let's start from the first `top_250`.

```{r libraries and data, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(atslib)
library(extrafont)
library(hrbrthemes)
theme_set(hrbrthemes::theme_ft_rc(base_family = "Ubuntu Condensed"))

top_250 <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-03/top_250.csv')

top_250
```

This dataset contains 250 NHL & WHA career leaders and records for goals. First, we're going to tidy our dataset a bit. What is the most interesting here is total goals really depends on the career length?

The information about years of career contains in `years` variable. It has a structure of `YYYY-YY`. First of all, let split it to `year_start` and `year_end`.

```{r tidy data, warning=FALSE}
tidy_250 <- top_250 %>% 
  mutate(year_start = str_split(years, "-", simplify = T)[,1],
         year_end = str_split(years, "-", simplify = T)[,2]) %>% 
  mutate_at(vars(year_start), ~as.numeric(.)) %>% 
  mutate(century = ifelse(year_start < 2000, 19, 20),
         cc = str_split(year_start, 
                        as.character(century),
                        simplify = T)[,2],
         cc = as.numeric(cc)) %>%
  mutate(year_end = case_when(
    year_end > cc ~ paste0(19, year_end),
    TRUE ~ paste0(20, year_end)
  )) %>% 
  mutate_at(vars(year_start, year_end), ~as.numeric(.)) %>% 
  # We need to edit millenials ))
  mutate(year_end = ifelse(year_end < year_start,
                           year_end + 100,
                           year_end)) %>% 
  select(player, total_goals, year_start, year_end) 

tidy_250
```

Great! Now we have a very tidy dataframe. However, if we are going to plot career length of all 250 players, it can be a total mess on the plot. Let's select only first 50 players using `dplyr::top_()`. 

```{r welcome tidy_50, warning=FALSE}
tidy_50 <- tidy_250 %>% 
  top_n(50, total_goals)
```

I prefer to use discrete color scale instead of continuous. So I'm going to convert `total_goals` into factor. I really like to do it interactively with a new [questionr](https://juba.github.io/questionr/reference/icut.html) package! So easy!

```{r top_50 cut, warning=FALSE}
## Cutting tidy_50$total_goals into tidy_50$total_goals_rec
tidy_50$total_goals_rec <- cut(tidy_50$total_goals,
                               include.lowest=TRUE,  right=TRUE,
                               breaks = c(400, 500, 600, 700, 800, 900),
                               labels = c("400-500",
                                          "500-600",
                                          "600-700",
                                          "700-800",
                                          "800-900"))
```

Okay, so everything is ready for a plot! Please, notice, that I've already set the `ggplot` theme in the first `r` chunk via `theme_set`. I prefer to use dark themes everywhere so for this post I'll also use a dark one. Bob Rudis ([hrbrmstr](https://twitter.com/hrbrmstr)) has created a really nice one in his [hrbrthemes](https://github.com/hrbrmstr/hrbrthemes).

```{r plot, fig.height=9, fig.width=6, message=FALSE, warning=FALSE}
library(ggalt)
library(paletteer)

tidy_50 %>% 
  mutate(len = year_end - year_start) %>% 
  mutate(player = fct_reorder(player, year_start)) %>%
  ggplot(aes(x = year_start,
             xend = year_end,
             y = player,
             group = player)) + 
  ggalt::geom_dumbbell(aes(color = total_goals_rec),
                size_x = 1.3,
                size_xend = 1.3,
                size = 1) +
  labs(x = "Years active", y = "",
       title = "NHL Career Leaders and Records for Goals",
       subtitle = "TOP 50",
       caption = "Made by @atsyplen as a #TidyTuesday contribution") +
  paletteer::scale_color_paletteer_d(name = "Total goals:",
                                     "wesanderson::Zissou1") +
  theme(panel.grid.major.y = element_blank())
```

Let's create a bit more meaningful plot. It's very interesting how `total_goals` are spread inside career lengths groups! For this one I'd prefer a boxlplot.

```{r boxplot, warning=FALSE}

tidy_250 %>% 
  mutate(len = year_end - year_start) %>% 
  mutate(len = cut(len,
                   include.lowest=TRUE,  right=TRUE,
                   breaks = c(5, 10, 15, 20, 25, 30, 35, 40),
                   labels = c(
                     "5-10",
                     "10-15",
                     "15-20",
                     "20-25",
                     "25-30",
                     "30-35",
                     "35-40"
                   ))) %>% 
  group_by(len) %>% 
  mutate(outlier.high = total_goals > quantile(total_goals, .75, na.rm = T) + 1.50*IQR(total_goals, na.rm = T),
         outlier.low = total_goals < quantile(total_goals, .25, na.rm = T) - 1.50*IQR(total_goals, na.rm = T),
         outlier.color = case_when(outlier.high ~ "red",
                                   outlier.low ~ "steelblue",
                                   outlier.low == F | outlier.high == F ~ "black")) %>%
  ggplot(aes(x = len, y = total_goals)) +
  stat_boxplot(geom ='errorbar',
               width = .25,
               color = "white") +
  geom_boxplot(outlier.shape = NA,
               color = "white",
               fill = "white",
               alpha = .3) +
  geom_jitter(aes(color = outlier.color),
              width = .1,
              alpha = .6,
              show.legend = F) +
  scale_y_continuous(breaks = seq(100, 900, 100),
                     labels = seq(100, 900, 100)) +
  ggrepel::geom_text_repel(data = . %>% group_by(len) %>% 
                             filter(total_goals == max(total_goals)),
                           aes(label = player), segment.colour = "white",
                           color = "white", family = "Ubuntu Condensed") +
  ggsci::scale_color_lancet() +
  labs(y = "Total goals scored in career",
       x = "Career length, years",
       title = "NHL Career Leaders and Records for Goals",
       subtitle = "TOP 250",
       caption = "Made by @atsyplen as a #TidyTuesday contribution") +
  theme(panel.grid.major.x = element_blank())

```

